---
title: "A Predictive Analysis of Coronary Heart Disease(CHD) Using Health and Lifestyle Variables"
date: "2025-03-28"
author: "Anteneh G. Yitayal 256983"
output:
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE,
message=FALSE,
tidy.opts=list(width.cutoff = 80),
tidy = FALSE)
```

## Introduction

1.  sex: Categorical variable indicating the gender of the individual (e.g., "Male" or "Female").
2.  age: Numeric variable representing the age of the individual (in years).
3.  education: Numeric variable representing the education level, possibly on a scale (e.g., 1 = low, 4 = high).
4.  smoker: Binary numeric variable indicating whether the individual is a smoker (1 = yes, 0 = no).
5.  cpd: Numeric variable representing cigarettes per day, relevant if the individual is a smoker.
6.  stroke: Binary numeric variable indicating whether the individual has had a stroke (1 = yes, 0 = no).
7.  HTN (Hypertension): Binary numeric variable indicating the presence of hypertension (1 = yes, 0 = no).
8.  diabetes: Binary numeric variable indicating whether the individual has diabetes (1 = yes, 0 = no).
9.  chol (Cholesterol): Numeric variable representing the individual's cholesterol level (in mg/dL).
10. DBP (Diastolic Blood Pressure): Numeric variable indicating the individual’s diastolic blood pressure (in mmHg).
11. BMI (Body Mass Index): Numeric variable representing the individual's BMI, which is a measure of body fat based on height and weight.
12. HR (Heart Rate): Numeric variable representing the individual's resting heart rate (in beats per minute).
13. CHD (Coronary Heart Disease): Categorical target variable indicating whether the individual has coronary heart disease ("Yes" or "No").

```{r, include=FALSE}
library(tidyverse)
if (!require(ROCR)) {
    install.packages(ROCR)
    library(ROCR)
}
```

The data set consists of 4,238 observations and 13 attributes related to cardiovascular health, specifically focusing on risk factors for Coronary Heart Disease (CHD). The attributes include demographic information such as sex and age, lifestyle factors like smoker and cigarettes per day (cpd), and clinical data such as hypertension (HTN), diabetes, cholesterol (chol), diastolic blood pressure (DBP), body mass index (BMI), and heart rate (HR). The data set also includes education level and a target variable CHD, indicating whether the individual has Coronary Heart Disease ("Yes" or "No"). This data set provides a comprehensive set of features that can be used for predictive modeling and analysis of factors influencing CHD risk.

```{r, include=FALSE}
#load the data set and view data
chd_df <- read_csv("chd.csv", show_col_types = FALSE)
data_size = dim(chd_df)
View(chd_df)
sprintf("Records: %d Features: %d", data_size[1], data_size[2])
```

The data set has a total of 204 missing values, with education level accounting for 51% of the missing records. This suggests that a significant portion of the missing data is concentrated in the education attribute. Addressing these missing values is crucial for ensuring the accuracy of any analysis, and potential strategies include imputation (replacing missing values with estimates) or removing rows with missing values, depending on the analysis approach.

```{r, include=FALSE}
#check missing values
check_missing_values <- function(data) {
  sprintf("#NaN values: %d", sum(is.na(data)))
  colSums(is.na(data))
}

check_missing_values(chd_df)
```

```{r}
data_df <- na.omit(chd_df)
check_missing_values(data_df)
dim(data_df)
```

```{r, include=FALSE}
# convert categorical into factor
data_df$sex <- as.factor(data_df$sex)
data_df$smoker <- as.factor(data_df$smoker)
data_df$stroke <- as.factor(data_df$stroke)
data_df$HTN <- as.factor(data_df$HTN)
data_df$diabetes <- as.factor(data_df$diabetes)
data_df$HTN <- as.factor(data_df$HTN)
data_df$education <- as.factor(data_df$education)

# Convert CHD to a binary factor (0 = No CHD, 1 = CHD)
data_df$CHD <- as.factor(data_df$CHD)

#check structure again
str(data_df)
```

```{r, include=FALSE}
summary(data_df)
```

```{r}
#check structure and compute Pearson's r correlation
str(data_df)
cor(dplyr::select(data_df, age, cpd, chol, DBP, BMI, HR))

```

About 85% of the data belongs to no chd, which is common in most medical datasets, such class imbalance ...

```{r}
# a high class imbalance between the target class
ggplot(data = data_df, aes(x = CHD)) +
  geom_bar(fill="skyblue", color="black") +
  labs(title="Distribution of Coronary Heart Disease(CHD)", x = "CHD", y ="Frequency")

prop.table(table(data_df$CHD))
```

```{r}
#our target variable is CHD, let's analyze it's relation with the other attributes
# proportion of chd between male and female
ggplot(data = data_df, aes(x = sex, fill = CHD)) + 
  geom_bar(color = "black") +
  labs(title = "Sex vs CHD",x = "Sex", y = "Frequency") +
  scale_fill_manual(values = c("No" = "skyblue", "Yes" = "tomato"))

prop.table(table(data_df$sex, data_df$CHD))
```

```{r}
str(data_df)
```

Most people start perimenopause in their late 40s. You officially reach menopause after you haven't had a period for one year. On average, Dr. Cole said people typically reach menopause between ages 51 and 52, but that varies from person to person.

```{r, fig.width=10, fig.height=4, fig.cap="Figure 1: your caption here.", fig.align="center"}
# seems like on average those who got chd are older than those who don't 
ggplot(data = data_df, aes(x = CHD, y = age)) +
  geom_boxplot(fill = "skyblue", color = "black") +
  labs(title = "Age by CHD", x = "CHD", y = "Age")

#age and sex make sense; which is normally high for male but as menopause starts it's getting worst for females
ggplot(data = data_df, aes(x = CHD, y = age, fill = sex)) +
  geom_boxplot(color = "black") +
  labs(title = "Age by CHD and Sex", x = "CHD", y = "Age") +
  scale_fill_manual(values = c("skyblue", "lightpink")) + 
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14),
    axis.title = element_text(size = 12)
  )

#biased towards general public (less college degree) so not appropriate to compare
ggplot(data = data_df, aes(x = education, fill = CHD)) + 
  geom_bar(color = "black") +
  labs(title = "Education Levels vs CHD",x = "Education Level", y = "Frequency") +
  scale_fill_manual(values = c("No" = "skyblue", "Yes" = "tomato"))

prop.table(table(data_df$education, data_df$CHD))

# smoking seems doesn't have an effect on getting chd based on the data
ggplot(data = data_df, aes(x = factor(smoker, levels = c(0, 1), labels = c("No", "Yes")), fill = CHD)) + 
  geom_bar(color = "black") +
  labs(title = "Smoker vs CHD",x = "Smoker", y = "Frequency") +
  scale_fill_manual(values = c("No" = "skyblue", "Yes" = "tomato"))

prop.table(table(data_df$smoker, data_df$CHD))

# stroke seems not an issue
ggplot(data = data_df, aes(x = factor(stroke, levels = c(0, 1), labels = c("No", "Yes")), fill = CHD)) + 
  geom_bar(color = "black") +
  labs(title = "Stroke vs CHD",x = "Stroke", y = "Frequency") +
  scale_fill_manual(values = c("No" = "skyblue", "Yes" = "tomato"))

prop.table(table(data_df$stroke, data_df$CHD))

#hypertension almost no effect
ggplot(data = data_df, aes(x = factor(HTN, levels = c(0, 1), labels = c("No", "Yes")), fill = CHD)) + 
  geom_bar(color = "black") +
  labs(title = "Hypertension vs CHD",x = "Hypertension", y = "Frequency") +
  scale_fill_manual(values = c("No" = "skyblue", "Yes" = "tomato"))

prop.table(table(data_df$HTN, data_df$CHD))

#not clear 
ggplot(data = data_df, aes(x = factor(diabetes, levels = c(0, 1), labels = c("No", "Yes")), fill = CHD)) + 
  geom_bar(color = "black") +
  labs(title = "Diabetes vs CHD",x = "Diabetes", y = "Frequency") +
  scale_fill_manual(values = c("No" = "skyblue", "Yes" = "tomato"))

prop.table(table(data_df$diabetes, data_df$CHD))
  

#people who developed chd tends to have higher cholesterol level on average and there are more extreme cholesterol in those who developed
ggplot(data = data_df, aes(x = CHD, y = chol)) +
  geom_boxplot(color = "black", fill="skyblue") +
  labs(title = "Cholestrol Level and CHD", x = "CHD", y = "Cholestrol Level") +
  scale_fill_manual(values = c("skyblue", "lightpink")) + 
  theme_minimal()

#similar trend as cholesterol level
ggplot(data = data_df, aes(x = CHD, y = DBP)) +
  geom_boxplot(color = "black", fill="skyblue") +
  labs(title = "Diastolic blood pressure and CHD", x = "CHD", y = "Diastolic blood pressure Level") +
  scale_fill_manual(values = c("skyblue", "lightpink")) + 
  theme_minimal()

#similar trend as cholesterol level
ggplot(data = data_df, aes(x = CHD, y = BMI)) +
  geom_boxplot(color = "black", fill="skyblue") +
  labs(title = "Body Mass Index and CHD", x = "CHD", y = "Body Mass Index") +
  scale_fill_manual(values = c("skyblue", "lightpink")) + 
  theme_minimal()

#similar trend as cholesterol level
ggplot(data = data_df, aes(x = CHD, y = HR)) +
  geom_boxplot(color = "black", fill="skyblue") +
  labs(title = "Heart Rate and CHD", x = "CHD", y = "Heart Rate") +
  scale_fill_manual(values = c("skyblue", "lightpink")) +
  theme_minimal()

ggplot(data = data_df, aes(x = CHD, y = cpd)) +
  geom_boxplot(color = "black", fill="skyblue") +
  labs(title = "Cigarettes per day and CHD", x = "CHD", y = "Cigarettes per day") +
  scale_fill_manual(values = c("skyblue", "lightpink")) +
  theme_minimal()


ggplot(data = data_df, aes(x = age, y = chol, color = CHD)) +
  geom_line(size = 1) +
  labs(title = "Yearly Values Trend", x = "Year", y = "Value") +
  theme_minimal()
```

```{r}
library(caret)

#set seed, and create train test split with stratified sampling
set.seed(123) 
train_size = 0.7
trainIndex <- createDataPartition(data_df$CHD, p = train_size, list = FALSE)

train_set <- data_df[trainIndex, ]
test_set <- data_df[-trainIndex, ]

# Check class distribution in train and test sets:
table(train_set$CHD) / nrow(train_set)
table(test_set$CHD) / nrow(test_set)

#str(train_set)

```

### Methods

```{r}
#fit with all features 
log_reg_all <- glm(CHD ~ .,
    data = train_set,
    family = binomial
)

summary(log_reg_all)
```

#### Logistic Regression

```{r}
#fit with significant features and the AIC: Drops so this is better than the model with all features
log_reg_sig <- glm(CHD ~ sex + age + cpd + HTN + diabetes + DBP,
    data = train_set,
    family = binomial
)

summary(log_reg_sig)
```
```{r}
#to see the effect of menopause, a new variable is create 1 if female and aged greater than 45, no significant for this data
log_reg_sig_menopause <- glm(CHD ~ age + age:sex + cpd + HTN + diabetes + DBP, 
             family = binomial, data = train_set)
summary(log_reg_sig_menopause)
```

```{r}
#compare the two models with AIC (goodness of fit + model complexity)
AIC(log_reg_all, log_reg_sig, log_reg_sig_menopause)
```

```{r}
# check the dummy variable for chd
contrasts(train_set$CHD)

#predict the test dataset
test_chd = test_set$CHD
glm.probs <- predict(log_reg_sig_menopause, newdata = test_set, type = "response")
glm.probs[1:10]

#set threshold and fill the placeholder
threshold = 0.5
glm.pred <- rep("No", nrow(test_set)) # A "placeholder" filled with as many No values as the number of observations (rows) in test_data
glm.pred[glm.probs > 0.5] <- "Yes" # replace with Yes values according to the glm.probs threshold

# confusion matrix
confusion_matrix <- table(glm.pred, test_chd)
confusion_matrix

# accuracy
mean(glm.pred == test_chd)
# error
mean(glm.pred != test_chd)
```

```{r}
# compute the prediction and plot ROC curve
predob <- ROCR::prediction(glm.probs, test_chd)
# then we need a "performance" object for roc and auc
perf <- ROCR::performance(predob, "tpr", "fpr")
auc_perf <- ROCR::performance(predob, measure = "auc")

auc_value = unlist(auc_perf@y.values)

# now the plot!
roc_data <- data.frame(
  fpr = unlist(perf@x.values),  # False Positive Rate
  tpr = unlist(perf@y.values)   # True Positive Rate
)

ggplot(roc_data, aes(x = fpr, y = tpr)) +
  geom_line(color = "blue", size = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray") +
  labs(
    title = paste("Logistic Regression ROC Curve (AUC =", round(auc_value, 3), ")"),
    x = "False Positive Rate (FPR)",
    y = "True Positive Rate (TPR)"
  ) +
  theme_minimal() + 
  theme(
    plot.title = element_text(hjust = 0.5, size = 14),
    axis.title = element_text(size = 12)
  )
```


#### KNN
```{r}

```

### Results and Discussion

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### Conclusion
